---
title: "R Notebook"
output: html_notebook
---

#Question 1 - Beta
```{r}
data <- read.csv("jpm_spx.csv", header = F)
vix <- read.csv("vix.csv", header = F)
colnames(data) <- c("DATE", "JPM", "SPX")
colnames(vix) <- c("DATE", "VIX")
data$VIX <- vix$VIX

data$DATE <- as.Date(data$DATE, format = "%Y-%m-%d")


jpm_spx <- lm(JPM ~ SPX, data)
plot(x = data$SPX, y = data$JPM, xlab = "SPX Returns", ylab = "JPM Returns", main = "JPM vs SPX Returns w/ Regression Line")
abline(jpm_spx)
text(x = 0.07, y = -.2, labels = eqn)

cc <- jpm_spx$coefficients
eqn <- paste("JPM =", paste(round(cc[1],2), paste(round(cc[-1],2), names(cc[-1]), sep=" * ", collapse=" + "), sep=" + "), "+ e")

install.packages("rollRegres")
library(rollRegres)

fits <- roll_regres(JPM ~ SPX, data, width = 63)
plot(fits$coefs[,2], type = "l", x = data$DATE, xlab = "Date", ylab = "Beta")


#VIX Betas
jpm_vix <- lm(JPM ~ VIX, data)
plot(x = data$VIX, y = data$JPM, xlab = "VIX Returns", ylab = "JPM Returns", main = "JPM vs VIX Returns w/ Regression Line")
abline(jpm_vix)
text(x = -0.15, y = -.2, labels = eqn2)

cc2 <- jpm_vix$coefficients
eqn2 <- paste("JPM =", paste(round(cc2[1],2), paste(round(cc2[-1],2), names(cc2[-1]), sep=" * ", collapse=" + "), sep=" + "), "+ e")

#Strong correlation with market - not a good factor. 
pairs(data[,-1])
cor(data$SPX, data$VIX)
```


#Question 2 - Performance Modelling
Isolating daily returns for the four year period from 2001 to 2004, I found:

Average Daily Return: 0.59%
Daily Standard Deviation: 1.54%

Annualized Return = 148.40
Annualized SD = 24.44%

Sharpe = 148.40 / 24.44 = 6.07

```{r}
ff_data <- read.csv("ff_data.csv", header = F)
colnames(ff_data) <- c("DATE", "MKTRF", "SMB", "HML", "RF", "UMD")

strat <- read.csv("strat_returns.csv", header = T)
strat$Date <- ff_data$DATE
strat$Total = as.numeric(gsub("[\\%,]", "", strat$Total))/100

#Using FF Market - RF for the CAPM:
#Here, we do not use Risk-Free as it is a self-financing portfolio in Sharpe
#Use RF to Calculate Excess Market Return
#Using SPX returns and annualised rf / 252 from Cripsy04

spx_rf <- read.csv("spx_rf.csv", header = F)
colnames(spx_rf) <- c("DATE", "SPX", "RF")
strat <- cbind(strat, spx_rf[,-1])
strat$MRKRF <- strat$SPX - strat$RF
strat <- cbind(strat, ff_data[,-1])

model_capm <- lm(Total ~ MRKRF, strat)
summary(model_capm)

model_ff <- lm(Total ~ MRKRF + UMD + HML + SMB, strat)
summary(model_ff)

ordered <- strat[order(strat$Total),]
plot(ordered$Total, main = "Ordered Returns", xlab = "", ylab = "Daily Return"); grid()

winners = sum(strat$Total > 0)/nrow(strat)
losers = sum(strat$Total < 0)/nrow(strat)
neutral = sum(strat$Total == 0)/nrow(strat)

median_winner <- median(strat$Total[(strat$Total > 0)])
median_loser <- median(strat$Total[(strat$Total < 0)])

```

There are far more winners than losers (66% vs 34%). Furthermore, the median of the winners is higher than that of the losers (1% vs -0.7%). The strategy thus gains a larger value more often, indicating a solid performance. 

#Question 3 - Performance Attribution
```{r}
strat_09 <- read.csv("strat_0509.csv", header = T)
spx_09 <- read.csv("spx_rf_0509.csv", header = F)
colnames(spx_09) <- c("DATE", "SPX", "RF")
strat_09 <- cbind(spx_09, strat_09[,-1])

annual_mean <- mean(strat_09$Grand.Total)*252
annual_sd <- sd(strat_09$Grand.Total)*sqrt(252)
sharp_total <- annual_mean/annual_sd

annual_mean_long <- mean(strat_09$Long)*252
annual_sd_long <- sd(strat_09$Long)*sqrt(252)
sharp_long <- annual_mean_long/annual_sd_long

annual_mean_short <- mean(strat_09$Short)*252
annual_sd_short <- sd(strat_09$Short)*sqrt(252)
sharp_short <- annual_mean_short/annual_sd_short

model_05 <- lm(Grand.Total ~ (SPX-RF), strat_09)
summary(model_05)



ordered_09 <- strat_09[order(strat_09$Grand.Total),]
plot(ordered_09$Grand.Total, main = "Ordered Returns", xlab = "", ylab = "Daily Return"); grid()

winners_09 = sum(strat_09$Grand.Total > 0)/nrow(strat_09)
losers_09 = sum(strat_09$Grand.Total < 0)/nrow(strat_09)
neutral_09 = sum(strat_09$Grand.Total == 0)/nrow(strat_09)

median_winner_09 <- median(strat_09$Grand.Total[(strat_09$Grand.Total > 0)])
median_loser_09 <- median(strat_09$Grand.Total[(strat_09$Grand.Total < 0)])
```

Total Return in 2006 was 81.42%. Manufacturing contributed 39.56% (SIC).

GICS - NA (34.50%). Next actual GIC was Information Technology with 12.27%. 

## Question 4
Reading in GICS Pivot Table:

```{r}


gics_data <- read.csv("gics_pivot.csv", header = T)
gics_data <- gics_data[-(nrow(gics_data)), ]
summary(gics_data[,-1])

```

## Question 5
```{r}
dj_vol <- read.csv("dj_vol.csv", header = F)
dj_vol <- dj_vol[,c(1,3,4,5,6,7,8,9,10,11)]
colnames(dj_vol) <- c("DATE", "MONTH_VOL", "THREE_VOL", "COUNT", "WEIGHT", "R","VAR1_21", "VAR0_21", "VAR1_63", "VAR0_63")

avg_cor1 <- (dj_vol$MONTH_VOL - dj_vol$VAR0_21) / (dj_vol$VAR1_21 - dj_vol$VAR0_21)
avg_cor3 <- (dj_vol$THREE_VOL - dj_vol$VAR0_63) / (dj_vol$VAR1_63 - dj_vol$VAR0_63)

max_1 <- max(avg_cor1)
ind_1 <- which(avg_cor1 == max_1)
max_3 <- max(avg_cor3)
ind_3 <- which(avg_cor3 == max_3)

dj_vol$AVG1 <- avg_cor1
dj_vol$AVG3 <- avg_cor3

dj_vol$DATE <- as.Date(dj_vol$DATE, "%Y-%m-%d")

plot(x = dj_vol$DATE, y = dj_vol$AVG1, type = "l", xlab = "", ylab = "Correlation", main = "1 Month Volatility Plot", ylim = c(0,1))
par(new = T)
lines(x = dj_vol$DATE, y = dj_vol$VAR0_21, col = "red")
lines(x = dj_vol$DATE, y = dj_vol$VAR1_21, col = "green")
lines(x = dj_vol$DATE, y = dj_vol$MONTH_VOL, col = "blue")


plot(x = dj_vol$DATE, y = dj_vol$AVG3, type = "l", xlab = "", ylab = "Correlation", main = "3 Month Volatility Plot", ylim = c(0,1))
par(new = T)
lines(x = dj_vol$DATE, y = dj_vol$VAR0_63, col = "red")
lines(x = dj_vol$DATE, y = dj_vol$VAR1_63, col = "green")
lines(x = dj_vol$DATE, y = dj_vol$THREE_VOL, col = "blue")
```
